---
title: "STAT 441 Lecture 9"
subtitle: "Multilevel Models"
format: revealjs
execute:
  cache: true
---


# Motivation

## Comparing Football Plays{.smaller}

- Which of these two touchdown plays is more impressive?
  * 86-yard touchdown pass from Justin Herbert to Ladd McConckey ([video link](https://www.youtube.com/watch?v=Hu9flYe2VrA))
  * 64-yard touchdown pass from Cooper Rush to KaVontae Turpin ([video link](https://www.youtube.com/watch?v=U0Zp_is-X18))

. . . 

- Plays share many similarities but important differences
  - Context: down & distance, time left, score
  - Time of pass: number of pass rushers, pocket integrity
  - Receiver's actions after the catch
  
. . .

- Today: expected points facilitated nuanced comparison
- Multilevel models to determine which passers generate most EP per attempt


# Expected Points

## Overivew

- Goal: estimate avg. number of points eventually scored by teams from similar situation

- EPA: diff. in post- and pre-play EP
  - $\textrm{EPA} > 0$: successful for offense
  - $\textrm{EPA} < 0$: unsuccessful for offense

- **nflfastR**'s EP model to predict next scoring event in half  
  - Touchdown (7), field goal (3), safety (2)
  - Opposing touchdown (-7), field goal (-3), and safety (-2)
  - No score (0)

## Expected Points

- Vector of next score probabilities given play features $\boldsymbol{\mathbf{z}}$: $\boldsymbol{\pi}(\boldsymbol{\mathbf{z}}) = (\pi_{\textrm{TD}}(\boldsymbol{\mathbf{z}}), \ldots, \pi_{\textrm{oppFG}}(\boldsymbol{\mathbf{z}}))$

- Estimated w/ regression tree ensemble using XGBoost

:::{.callout-note icon=false}
## Definition: Expected Points
Given a game state feature vector $\boldsymbol{\mathbf{z}}$ and vector of drive outcome probabilities $\boldsymbol{\pi}(\boldsymbol{\mathbf{z}}),$ the *expected points* $\textrm{EP}(\boldsymbol{\mathbf{z}})$ is 
$$
\begin{align}
\textrm{EP}(\boldsymbol{\mathbf{z}}) &=  7\times\pi_{\textrm{TD}}(\boldsymbol{\mathbf{z}}) + 
3\times\pi_{\textrm{FG}}(\boldsymbol{\mathbf{z}}) + 
2\times\pi_{\textrm{SAF}}(\boldsymbol{\mathbf{z}}) \\
~&~~-2\times\pi_{\textrm{oSAF}}(\boldsymbol{\mathbf{z}}) - 
3\times\pi_{\textrm{oFG}}(\boldsymbol{\mathbf{z}})
- 7\times\pi_{\textrm{oTD}}(\boldsymbol{\mathbf{z}})
\end{align}
$$
:::

## League-Average 


- EPA on a future pass if you don't know anything else?
- Is the league average `mean(pbp2024$epa)` reasonable?


## Player-Specific Model

- Would your prediction change if you knew passer identity?
  - If Patrick Mahomes was throwing the pass, expect higher EPA
  - If Daniel Jones was throwing the pass, expect lower EPA

. . .

- $I$: number of unique passers in data set
- $Y_{ij}$: EPA on pass $j$ thrown by player $i$
- $\alpha_{i}$: avg. EPA/pass for player $i$ (unknown)
- Model: $Y_{ij} = \alpha_{i} + \epsilon_{ij}$

## Estimating $\alpha_{i}$ {.smaller}
- Idea 1: group by `passer_player_id` & compute `mean(epa)` for each player
- Equivalent (and more extensible): fit a linear model w/ least squares
  - Regress `epa` on *categorical* `passer_player_id`
  - Must convert `passer_player_id` to a `factor` variable
  - Useful to manually set reference player (e.g., Aaron Rodgers)

- Estimates $\beta_{0} = \alpha_{\textrm{Rodgers}}$ and $\beta_{i} = \alpha_{i} - \alpha_{\textrm{Rodgers}}$


## Partial Pooling{.smaller}

- If $i$ threw many passes, $\hat{\alpha}_{i}$ accurately estimates latent ability $\alpha_{i}$
  - Prefer to use $\hat{\alpha}_{i}$ over global mean $\overline{y}$
  
- If $i$ threw very few passes, $\hat{\alpha}_{i}$ can be very noisy
  - Global mean $\overline{y}$ arguably better than $\hat{\alpha}_{i}$
  
- What about for players b/w extremes?

. . .

- Idea: $w_{i} \times \hat{\alpha}_{i} + (1 - w_{i}) \times \overline{y}$
  - If $n_{i}$ very large, $w_{i}$ should be closer to 1
  - If $n_{i}$ very small, $w_{i}$ should be closer to 0
- Problem: how to specify weights *in data-driven way*?

# Multilevel Models

## Model Specification{.smaller}
- Level 1: observed EPA for player $i$ normally distributed around $\alpha_{i}$
- Level 2: Latent player abilities $\alpha_{i}$'s are themselves normally distributed around $\mu$
$$
\begin{align}
\textrm{Level 1}&: &\quad  Y_{ij} &= \alpha_{i} + \epsilon_{ij}; \epsilon_{ij} \sim N(0, \sigma^{2}) \quad \textrm{for all}\ j = 1, \ldots, n_{i},\ i = 1, \ldots, I \\
\textrm{Level 2}&: &\quad \alpha_{i} &= \alpha_{0} + u_{i}; u_{i} \sim N(0, \sigma^{2}_{\alpha}) \quad \textrm{for all}\ i = 1, \ldots, I
\end{align}
$$

- $\alpha_{0}$: average per-pass EPA over super-population of passers
- $\sigma$ captures "within-player" variability in EPA pass-to-pass
- $\sigma_{\alpha}$ captures "between-player" variability in per-pass EPA

. . .

- Level 2 responses $\alpha_{i}$ are **not observable**
- "Borrowing strength": predition for player $i'$ informed by their data *and everyone else's data*
  - Estimate of $\alpha_{i}$ informed by $i$'s data **and $\alpha_{0}$
  - Estimates of $\alpha_{i}$ determine estimate of $\alpha_{0}$
  
## Hierarchical Structure

- Often data exhibits hierarchical grouping structure
  - Passes group by QB, players in teams etc.
- Grouping variable may be relevant to outcome & induces correlation
- Outcomes from same group may be tightly clustered around group average

## Fitting Multilevel Models

- The `(1 | passer_player_id)` tells `lmer()` to include a *random intercept* for each passer

```{r}
#| label: fit-random-intercept
#| echo: true
#| eval: false
#| message: false
library(lme4)
multilevel_fit <-
  lmer(formula = epa ~ 1 + (1|passer_player_id), 
       data = pass2024)
```

## Extracting Estimates{.smaller}
- Use `ranef()` to extract estimates $\hat{u}_{i}$'s
- Use `coef` to get $\hat{\alpha}_{i} = \hat{\alpha}_{0} + \hat{u}_{i}$
  - Only valid when there is one grouping variable
  - W/ more grouping variables, random intercepts not identified
- `coef()` and `ranef()` return lists w/ one element per grouping variable
  - Each list element is a data table

```{r}
#| label: display-coef
#| eval: false
#| echo: true
tmp <- coef(multilevel_fit)
```


## Adjusting For Additional Covariates

- Simple model doesn't account for formation, whether QB was hit while throwing, etc.
- $\boldsymbol{\mathbf{x}}$ contains: `air_yards`, `shotgun`, `qb_hit`, `pass_location`, `posteam_type`
- Effects of these factors are constant (i.e., **fixed**) across all passers
$$
\begin{align}
\textrm{Level 1}&: &\quad  Y_{ij} &= \alpha_{i} + \boldsymbol{\mathbf{x}}_{ij}^{\top}\boldsymbol{\beta} +  \epsilon_{ij}; \quad \epsilon_{ij} \sim N(0, \sigma^{2}) \\
\textrm{Level 2}&: &\quad \alpha_{i} &= \alpha_{0} + u_{i}; \quad u_{i} \sim N(0, \sigma^{2}_{\alpha}) 
\end{align}
$$

## Fitting Our Elaborate Model{.scrollable}

```{r}
#| label: multilevel-full
#| eval: false
#| echo: true
ml_fit_full <-
  lme4::lmer(formula = epa ~ 1 + (1|passer_player_id) + 
               air_yards + posteam_type + shotgun +
               no_huddle + qb_hit + pass_location, data = pass2024)
```



## Looking Ahead

- EPA is based on starting & ending game state
- 2 phases of passing play: ball in air + after the catch
- Currently analysis implicitly credits QBs for both phases

- [Lecture 10](../../lectures/lecture10.qmd): divide total EPA among relevant offensive players
- Develop a version of WAR
  - Multilevel model to estimate per-play contribution
  - Scale model estimates by opportunities to create replacement-level shadow